##Activity Analysis


```{r setup, message = FALSE, warning = FALSE}
library(dplyr)
library(xtable)
library(ggplot2)
```

####This code is used to import the data set and convert the class of data in the date column to type Date.
####Assumes all of the necessary files are in your working directory
```{r loadData}

activity <- read.csv('activity.csv', header = TRUE, stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date, '%Y-%m-%d')

```


####Next we are going to determine the number of steps taken per day and calculate the mean and median number of steps per day.
####Note that NA values are excluded.
```{r FirstActivitySummary}
#Remove NA data
activity1 <- activity[!is.na(activity$steps),]

#Group the data by date and sum the number of steps for each day
activitySummary <- activity1 %>% group_by(date) %>% summarise(NumSteps = sum(steps), StepMean = mean(steps), StepMedian = median(steps))


```

####Then we show the total steps, mean, and median steps per day
```{r showFirst, results ='asis', warning = FALSE}
xt <- xtable(activitySummary)
print(xt, type = 'html')
```


####Building on the previous step, we now show the results of the activity summary and plot a histogram of the number of steps per day
```{r activitySummaryplot, fig.path = getwd(), fig.show = 'asis'}

hist(activitySummary$NumSteps, xlab = 'Number of steps per day', ylab = '# of Days in Range' , main = '')

```

***

####The average number of steps in each interval is shown in the graph below
```{r intervalAverage}

intervalSummary <- activity1 %>% group_by(interval) %>% summarise(StepMean = mean(steps))

ggplot(aes(x = interval, y = StepMean), data = intervalSummary) + geom_line() + labs(x = 'Interval', y = 'Average Number of Steps')

maxStepinterval <- intervalSummary$interval[which(intervalSummary$StepMean == max(intervalSummary$StepMean))]

```

####The 5 minute interval containing the highest average number of steps is the `r maxStepinterval`th


***

###In this part of the analysis we'll introduce the missing values back into the data.

####First we determine how many rows have missing NA step data
```{r missingVals}
numMissing <- nrow(activity[is.na(activity$steps),])
```

####The number of missing values in the data set is `r numMissing`
 
####Then we will fill the missing values with the mean step count for that interval
```{r fillMissing}

activityFull <- activity

for(i in 1:nrow(activityFull)){
  
  if( is.na(activityFull$steps[i]) ){
    intervalIndex <- which(intervalSummary$interval == activityFull$interval[i])
    activityFull$steps[i] <- intervalSummary$StepMean[intervalIndex]
  }
}

#Calculate the new total step count, mean, and median per day
activitySummaryFull <- activityFull %>% group_by(date) %>% summarise(NumSteps = sum(steps), StepMean = mean(steps), StepMedian = median(steps))

```



```{r activitySummarwithNAs, fig.path = getwd(), fig.show = 'asis'}

hist(activitySummaryFull$NumSteps, xlab = 'Number of steps per day', ylab = '# of Days in Range' , main = '')
```


####Looking at the above plot, it is very similar to the first, with the notable exception that the number of days that fall in the 10,000 - 15,000 step range is greater. By imputing the missing data we've increased the estimate of total daily steps.


###Weekdays vs Weekends

```{r daysofWeek, results = 'hide'}
activityFull$dateName <- weekdays(activityFull$date)

for(i in 1:length(activityFull$dateName)){
  
  if(activityFull$dateName[i] == 'Saturday' | activityFull$dateName[i] == 'Sunday'){
    activityFull$dateName[i] = 'Weekend'
  }
  else{
    activityFull$dateName[i] = 'Weekday'
  }
}

#Makes a two level factor column within activityFull
factor(activityFull$dateName)

intervalSummary <- activityFull %>% group_by(interval,dateName) %>% summarise(StepMean = mean(steps, na.rm = TRUE))


```


###Plot the average steps by interval, broken up by weekends and weekdays
```{r daysPlot, fig.path = getwd(), fig.show = 'asis'}

ggplot(aes(x = interval, y = StepMean), data = intervalSummary) + geom_line() + labs(x = 'Interval', y = 'Average Number of Steps') + facet_grid(dateName~.)

```

